---
title: 'sfnetworks: Tidy Geospatial Networks in R - OpenGeoHub Summer School 2021'
author: "Andrea Gilardi"
date: "01/09/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, fig.align = "center")
```

## Introduction

First, please note that we will present the development version of `sfnetworks` that can be installed as follows: 

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("luukvdmeer/sfnetworks", "develop")
```

Load some packages, data and options: 

```{r, message=FALSE}
Sys.setenv(`_R_S3_METHOD_REGISTRATION_NOTE_OVERWRITES_` = "false")
library(sf)
library(tidygraph)
library(sfnetworks)
library(ggplot2)
library(tmap)
library(magrittr)

iow <- st_read("https://github.com/sfnetworks/OGH2021/raw/master/isle-of-wight.gpkg", quiet = TRUE)
```

The object `iow` represents a subset of the street segments that define the road network of the Isle of Wight, an island located in the south of England, according to a driving mode of transport. 

Let's see a plot: 

```{r}
par(mar = rep(0, 4))
plot(iow)
```

We can already notice that there are several groups of segments that look disconnected from the main network.

## From LINESTRING(s) to sfnetwork object

Now we can convert the "raw data" into a sfnetwork object.

```{r}
sfn <- as_sfnetwork(iow, directed = FALSE)
```

Let's see the printing method

```{r}
sfn
```

We can notice that:

1. The output reports a short description of the sfnetwork object summarising, among the other things, the number of nodes and edges, the CRS, the number of components.
2. We know that the sfnetwork class inherits the tbl_graph class (see `?tbl_graph`) which means that we can apply all tidygraph and igraph methods to sfnetwork objects (we will see several examples later). Moreover, we can see that, by default, the nodes represent the active geometry (more details later on).
3. The connectivity among the road segments is summarised using two columns in the edge table named `"from"` and `"to"`.

Moreover:

1. The nodes of the sfnetwork object are created at the (unique) endpoints of the input lines. This has importance consequences (see below).
2. The argument `directed = FALSE` forces the construction of an undirected network. See the introductory vignette for examples on how to create an sfnetwork object mixing one-way and two-way roads.
3. The are several other ways to create an sfnetwork object. See `?sfnetwork` and `?as_sfnetwork` for more examples.
4. We can adjust the number of features printed for the active and inactive tables using the options `sfn_max_print_active` and `sfn_max_print_inactive`. For example:

```{r}
options(sfn_max_print_active = 2L, sfn_max_print_inactive = 1L)
sfn
```

Warning: this is still an experimental feature that exist only in the develop branch.

Let's see a plot again

```{r}
par(mar = rep(0, 4))
plot(sfn)
```

The sfnetwork method for the plot function is a wrapper around `plot.sfc_GEOMETRY`, which means that we can adjust several aspects of the plot: 

```{r}
par(mar = rep(0, 4))
plot(sfn, cex = 0.7, lty = 2, pch = 17)
```

There is also an autoplot method

```{r}
autoplot(sfn) + theme_minimal()
```

We will see more complex examples later on. 

## Active and inactive geometries

A sfnetwork object is a multitable object composed by two sf tables (one for the nodes and one for the edges). There is always one of the two tables which is labeled as "active". That table is the target for the data manipulation verbs.

For example

```{r}
options(sfn_max_print_active = 0L, sfn_max_print_inactive = 0L)
sfn
```

You can switch the active table using the function "activate" or the shortcuts `%N>%` and `%E>%` (just for pipe chains in interactive coding sessions)

```{r}
activate(sfn, edges)
```

## Pre-processing steps

As already mentioned, real-world spatial networks may require a bunch of pre-processing steps before we can apply sf or igraph algorithms. Hence, we implemented a collection of morphers (see below for more details) to process the input data:

```{r}
sfn_clean <- sfn %E>%
  convert(to_spatial_subdivision) %>%
  arrange(edge_length()) %>%
  convert(to_spatial_simple) %>%
  convert(to_spatial_smooth) %>%
  convert(to_components) %>%
  convert(to_spatial_simple, .clean = TRUE)
```

Check the companion slides and the package's vignettes for several toy examples showing the importance of these morphers.

Let's check the printing

```{r}
sfn_clean
```

and a plot 

```{r}
plot(sfn_clean)
```

A few comments:

1. The spatial and graph morphers (i.e. to_spatial_subdivision, to_spatial_simple, to_spatial_smooth, and to_components) are used to create temporary alternate representation of the input network. On the other hand, the function `convert()` is used to extract this temporary representation (or its first graph in case of multiple representations) and convert it into an sfnetwork object. See the fifth package vignette for more details and check `?convert`.
2. The "clean" version of the network is much simpler than the original one since we extracted only the nodes and edges that belong to the first component of the network. We can clearly notice that from the printing and plot methods.

## Computing graph measures

We can use some dplyr verbs (which are loaded by tidygraph) to modify the edges and nodes tables. For example, the following code computes a simple graph measure (e.g. vertex degree) and adds it to the corresponding tables:

```{r}
sfn_clean <- sfn_clean %N>%
  mutate(degree = centrality_degree())
```

We can add a column in the edges table measuring the length of each LINESTRING object:

```{r}
sfn_clean <- sfn_clean %E>%
  mutate(weight = as.numeric(edge_length()))
```

and compute weighted graph measures:

```{r}
sfn_clean <- sfn_clean %>%
  mutate(betw = centrality_edge_betweenness(weights = weight, directed = FALSE))
```

Let's see a print

```{r}
options(sfn_max_print_active = 2L, sfn_max_print_inactive = 2L)
sfn_clean
```

and some (more elaborate) plots:

```{r}
ggplot() +
  geom_sf(data = st_as_sf(sfn_clean, "edges"), col = "grey", size = 1) +
  geom_sf(data = st_as_sf(sfn_clean, "nodes"), aes(col = degree, size = degree)) +
  theme_minimal() +
  scale_color_viridis_c() +
  scale_size_continuous(range = c(1.5, 3)) +
  guides(color = guide_legend(), size = guide_legend()) +
  labs(col = "Node degree", size = "Node degree") +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(3, "lines"),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 10)
  )
```

or

```{r}
ggplot() +
  geom_sf(data = st_as_sf(sfn_clean, "edges"), aes(col = betw, fill = betw), size = 1.2) +
  theme_minimal() +
  scale_color_viridis_c(trans = "sqrt") +
  scale_fill_viridis_c(trans = "sqrt") +
  labs(col = "Edge centrality \nbetweenness", fill = "Edge centrality \nbetweenness") +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(4, "lines"),
    legend.text.align = 0
  )
```

More comments:

1. A graph measure can only be calculated if the correct table is marked as active, i.e. the following code returns an error

```{r, error = TRUE}
sfn_clean %>%
  activate(nodes) %>%
  mutate(betw = centrality_edge_betweenness())
```

2. The tidygraph package implements tens of functions for computing graph measures. See https://tidygraph.data-imaginist.com/reference/index.html for more details.
3. Not all dplyr verbs can be directly applied to a sfnetwork since we cannot tamper with the relational data structure.
4. Some tidygraph functions require you to always specify the arguments `"weights"` and `"directed"` (which are not inferred from the sfnetwork).
5. The node and edges tables can be extracted using the function `st_as_sf()` specifying the parameters `"nodes"` or `"edges"`. This is quite important for plotting routines. The same structure holds for `st_geometry` function.
6. The coordinates of the active geom can be extracted with st_coordinates:

```{r}
sfn_clean %N>% st_coordinates() %>% head()
```

or

```{r}
sfn_clean %E>% st_coordinates() %>% head()
```

7. The estimation of graph measures may be quite relevant for developing statistical models at the network level. See an example at the end of this presentation.

## Spatial filters (and other spatial operations)

A sfnetwork object is usually represented as two sf tables. Hence, we extended several sf functions defining a method for sfnetwork objects. For example, we can change the CRS of both tables as follows: 

```{r}
sfn_clean <- st_transform(sfn_clean, 27700)
```

This is quite relevant for all those packages (e.g. spatstat) that do not handle non-projected coordinates.

We can use `st_filter` to apply a binary spatial geometry predicate (such as `st_intersects`, `st_contains`, `st_is_within_distance` and so on). For example, we can compute a 5km circular buffer around the centroid of the network

```{r}
sfn_centroid <- st_centroid(st_as_sfc(st_network_bbox(sfn_clean)))
sfn_buffer <- st_buffer(sfn_centroid, 5000)
```

and filter only the nodes that intersect that buffer

```{r}
sfn_small <- sfn_clean %N>% st_filter(sfn_buffer)
```

Let's see a plot:

```{r}
par(mar = c(2.5, 2.5, 1, 1))
plot(sfn_clean, graticule = TRUE, axes = TRUE)
plot(sfn_buffer, add = TRUE, border = "orange", lwd = 3)
plot(sfn_small, add = TRUE, col = "red")
```

More details:

1. The function `st_network_bbox()` is used to compute the bounding box of edges' and nodes' bounding boxes (since they may be different, see `?st_network_bbox`).
2. The function st_filter is always applied to the active element of the sfnetwork object.
3. The spatial functions that may change type, shape or position of the geometries (e.g. st_union or st_jitter) are not directly supported. You can apply them after extracting the spatial table.
4. Please note that when nodes are removed, their incident edges are removed as well. However, when edges are removed, the nodes at their endpoints remain, even if they don’t have any other incident edges. For example:

```{r}
sfn_clean %E>% st_filter(sfn_buffer) %>% plot(lwd = 3)
plot(sfn_buffer, add = TRUE, border = "orange", lwd = 3)
```

Tips: 

1. The following code may be used to simulate an sfnetwork object where
a) `n` nodes are randomly placed in a unit square; 
b) two nodes are connected if their are closer than parameter `radius`. 
  
```{r}
set.seed(123)
my_random_nodes <- play_geometry(200, radius = 0.1)
my_random_sfn <- as_sfnetwork(my_random_nodes, coords = c("x", "y"))
```

Check the plot: 

```{r}
unit_square <- st_sfc(st_polygon(list(rbind(c(0, 0), c(0, 1), c(1, 1), c(1, 0), c(0, 0)))))
par(mar = rep(0, 4))
plot(unit_square, lwd = 2)
plot(my_random_sfn, add = TRUE)
```

2. If you need to subset a large `sfnetwork` object, the following approach may be more efficient thant the usual `st_filter` (although a bit more verbose). First, simulate some data: 

```{r}
set.seed(123)
my_random_nodes <- play_geometry(1e6, radius = 5e-4)
my_random_sfn <- as_sfnetwork(my_random_nodes, coords = c("x", "y"))
```

Define a polygon contained into the unit square: 

```{r}
small_square <- st_sfc(st_polygon(list(rbind(c(0.4, 0.4), c(0.4, 0.6), c(0.6, 0.6), c(0.6, 0.4), c(0.4, 0.4)))))
```

and filter only the nodes that are contained into that small square: 

```{r}
system.time({
  idxs <- st_contains(small_square, my_random_sfn %>% activate(nodes))
  my_small_random_sfn_v1 <- my_random_sfn %N>% slice(unlist(idxs))
})
```

The equivalent code with st_filter takes a little bit longer with equivalent results: 

```{r}
system.time({
  my_small_random_sfn_v2 <- my_random_sfn %N>% st_filter(small_square)
})
```

As always, let's plot the result (but, first, we need to repeat the test with less points): 

```{r}
my_random_nodes <- play_geometry(5e4, radius = 5e-3)
my_random_sfn <- as_sfnetwork(my_random_nodes, coords = c("x", "y"))

idxs <- st_contains(small_square, my_random_sfn %>% activate(nodes))
my_small_random_sfn <- my_random_sfn %N>% slice(unlist(idxs))

par(mar = rep(0, 4))
plot(unit_square, lwd = 2)
plot(small_square, lwd = 2, border = "darkred", add = TRUE)
plot(my_random_sfn, add = TRUE, pch = 46, draw_lines = FALSE, cex = 0.01)
plot(my_small_random_sfn, add = TRUE, pch = 46, draw_lines = FALSE, cex = 0.01, col = "darkred")
```

Clean ws

```{r}
rm(unit_square, small_square, my_random_nodes, my_random_sfn, my_small_random_sfn, my_small_random_sfn_v1, my_small_random_sfn_v2, idsx)
```

## Shortest paths 

Calculating shortest paths between pairs of nodes is a core task in network analysis. The sfnetworks package offers wrappers around the path calculation functions of igraph, making it easier to use them when working with spatial data and tidyverse packages.

Let's see an example. The following code can be used to estimate the path connecting the two random points in the Isle of Wight:

```{r}
set.seed(1)
iow_poly <- st_as_sfc(st_network_bbox(sfn_clean))
start_point <- st_sample(iow_poly, 1)
end_point <- st_sample(iow_poly, 1)
sfn_path <- st_network_paths(
  x = sfn_clean,
  from = start_point,
  to = end_point
)
sfn_path
```

The result is returned as an object of class tibble with one row per path. The columns node_paths and edge_paths the node and edge indices composing the path. We can plot the path as follows:

```{r}
edges_path <- sfn_clean %>%
  st_geometry("edges") %>%
  extract(do.call(c, sfn_path$edge_paths))

par(mar = rep(0, 4))
plot(sfn_clean, col = "grey")
plot(edges_path, add = TRUE, col = "#8da0cb", lwd = 4)
plot(start_point, col = "#66c2a5", pch = 20, add = TRUE, cex = 2)
plot(end_point, col = "#fc8d62", pch = 20, add = TRUE, cex = 2)
```

An analogous result can be obtained using the to_spatial_shortest_paths morpher:

```{r}
sfn_shortest <- sfn_clean %>%
  convert(to_spatial_shortest_paths, from = start_point, to = end_point)
```

Again, a plot

```{r}
par(mar = rep(0, 4))
plot(sfn_clean, col = "grey")
plot(sfn_shortest, add = TRUE)
plot(start_point, col = "#66c2a5", pch = 20, add = TRUE, cex = 2)
plot(end_point, col = "#fc8d62", pch = 20, add = TRUE, cex = 2)
```

The to_spatial_shortest_paths morpher internally calls `st_network_paths()` and returns an sfnetwork object that contains only the nodes/edges that belong to the shortest path.

More details:
1. If there is a column in the edges table named weight (such as in this case), the values in this column are automatically used as edge weights.
2. If there is no column named weight in the edges table, then `st_network_paths()` internally calculates the geographic lengths of the edges and uses those as weights.

Check the introductory vignettes for many many more examples (e.g. one-to-many routing, distance matrices, closest facility analysis, TSP, isochrones). 



